Co-evolution of lexical meaning \& pragmatic use

### 2018-03-18 paper/cogsci-submission2.tex
paper edits whole paper

This section engages in a formal thought experiment to address the question: if a population of
language users could freely combine different lexica with different dispositions to act on
them, what are conditions under which the majority view of scalar implicatures could have
evolved? --- Recall that the majority view is that scalar implicatures are non-lexicalized
pragmatic enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may}
are semantically weak expressions for which logically stronger expressions are salient, e.g.,
\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em
  all}. Still, the use of a less informative expression (\emph{some}) when a more informative
one (\emph{all}) could have been used can license a defeasible inference (\emph{some but not
  all}) that stronger alternative is false (cf. \citealt{horn:1972,gazdar:1979}). According to
the majority view, this is a pragmatic inference, not part of the conventional meaning.

In the following we consider what are perhaps two of the simplest non-trivial setups that speak
to this matter and reflect on their limitations in Section~\ref{sec:discussion}. Section
\ref{subsec:example} first investigates a minimal example intended to isolate effects of
replication and mutation and to gain insights into the sense in which semantic meaning and
pragmatic use can be said to co-evolve. We then turn to a larger and more realistic setup in
Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
for this case and their results.

@@

This section illustrates how replication and mutation act on the evolution of semantic meaning
and pragmatic use in the simplest non-trivial type space, which consists of only the four types
from Section~\ref{sec:languages+use}: a type has either lexicon $\Llack$ or $\Lbound$ and is
either a literal (level-$0$) or a pragmatic (level-$1$) language user. We may think of such a
population as a point in a two-dimensional space. The grey dotted square in
Fig.~\ref{fig:sketch} outlines this space. The space's first dimension determines the
proportion of types that lexicalize ambiguous $\msome$ (the $y$-axis in
Fig.~\ref{fig:sketch}). The second dimension determines the proportion of pragmatic language
users (the $x$-axis in Fig.~\ref{fig:sketch}). Populations at the corners of this space are
monomorphic; all agents in them are of a single type. Points away from the corners represent
mixed populations of various types.\footnote{This two-dimensional representation underspecifies
  the exact proportions of types found in a mixed population. The space's center, for instance,
  could be composed in equal parts of pragmatic $\Llack$ and of literal $\Lbound$. But it could
  equally well be a population that has all four types in equal proportions. This level of
  granularity will nevertheless be sufficient for our purposes, as we are only interested in
  sketching rough trajectories after replicator or mutator steps.} We can predict general
trends in evolutionary trajectories in this space based on what we know from the literature and
the setup from Section~\ref{sec:fitn--learn}, even before any further simulation results.

@@

Despite its simplicity, the diagram in Figure~\ref{fig:sketch} makes clear the sense in which
semantics and pragmatics might co-evolve. Pragmatic language use will not evolve under a strong
and fully expressive semantics (bottom part of Figure~\ref{fig:sketch}). A weak semantic
convention will not evolve without disposition towards pragmatic enrichments (left part of
Figure~\ref{fig:sketch}). But together a weak semantics and a pragmatic disposition for
enrichment can co-evolve. These abstract theoretical considerations also suggest that we need
both fitness based selection of communicative efficiency and pressure towards learnability to
see underspecified semantics and pragmatic language use emerge together.

@@

Fig.~\ref{fig:quiver} shows actual evolutionary trajectories in this type space. The replicator
steps shown in the top row of Fig.~\ref{fig:quiver} make the functional disadvantage of
$\Llack$ against $\Lbound$ apparent, particularly for its literal users and/or low
$\lambda$. They also highlight that the functional difference between literal and pragmatic
$\Lbound$ is small. Under functional pressure only, populations move toward the lower part of
the square but the steps become small as the lower edge, representing populations composed only
of $\Llack$ users, are reached. As for the mutator diagrams (bottom row of
Fig.~\ref{fig:quiver}), there are three main things to note. First, as expected from the
literature, a small bias for a type will lead to larger proportions to adopt it under $l=1$
(see lower left space in Figure~\ref{fig:quiver}). Second, if $l>0$, the dynamic is influenced
heavily also by asymmetries in production likelihoods. Third, these asymmetries in the
production likelihoods are in turn influenced by $\lambda$, so that $\lambda$ also has an
effect on iterated Bayesian learning (compare the middle and right-most diagram in the bottom
row of Fig.~\ref{fig:quiver}).

Fig.~\ref{fig:sketch} and \ref{fig:quiver} investigate replicator and mutator steps in
isolation. It remains to investigate how exactly these forces shape a population over time when
they apply together. This will also hinge on which types are present because the space of all
types determines whether production likelihoods asymmetrically favor certain types over others
(when $l>1$; see Appendix~\ref{sec:example-iter-bayes}). For this reason, we turn to a larger
setup next with a larger and more realistic set of types. Additionally, we want to explore
the effects of a learning bias in favor of simple lexical representations. One way of
motivating and formalizing such a bias is introduced in Section~\ref{sec:an-induct-learn}.

@@

A lexicon $L$ is a mapping $\Messgs \rightarrow R$ from messages to representations. With three
messages there are $6^3 = 216$ possible lexica. Some assign the same representations to more
than one message and others lexicalize the same representations but associate them with
different messages. Out of these possible lexica, three kinds are of particular
relevance. First, lexica that assign the same lexical representations to more than one
message. Such lexica lack in expressivity but may be favored by particular learning biases
nonetheless (see below). Second, lexica that conventionalize upper-bounds to realize a
one-to-one mapping of messages to states. Finally, lexica that do not lexicalize an upper
bound but allow it to be conveyed pragmatically due to the presence of a stronger lexical
item. There are six lexica of the second kind and six of the third.  The following three lexica
exemplify each kind:
\begin{align*}
  \begin{blockarray}{lccc}
    & & \underline{\Lall} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
     \mygray{\snone}  & 0 & 0 & 0\\
     \mygray{\ssome}  & 0 & 0 & 0\\
    \mygray{\sall}   & 1 & 1 & 1\\
    \end{block}
% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
  \end{blockarray} & &
 \begin{blockarray}{lccc}
    & & \underline{\Lbound} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 0 & 1\\
    \end{block}
%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
  \end{blockarray} & &
  \begin{blockarray}{lccc}
    & & \underline{\Llack} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 1 & 1\\
    \end{block}
%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
  \end{blockarray}
\end{align*}
Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
be equally likely; no information about the observed world state will be conveyed
by its users. In contrast, users of $\Lbound$ can communicate world states perfectly, no matter
whether they are literal or pragmatic users. Users of $\Llack$ can also communicate information
about the actual world state but need pragmatic language use to approximate a one-to-one
mapping between message use and states (see Section~\ref{sec:languages+use}).

@@

Note that while
different types may lexicalize the same representations, they may nevertheless map different
states to different overt messages. More informally, they speak different languages that
lexicalize the same concepts. Consequently, more often than not different types of the same
kind fail to understand each other.

@@

A complexity measure for lexical representations from Table~\ref{tab:concepts} is used to define a
learning bias that favors simpler representations over more complex ones. The prior probability of a
type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
complexity of the lexical representations in its image set. Lexica with simpler representations accordingly have a
higher prior. One simple way of defining such priors over lexica (and thereby types) is:
\begin{align*}
  P(L)  & \propto \prod_{r \in Im(L)} P(r)   \ \text{, with} & 
  P(r) & \propto \max_{r'}Compl(r') - Compl(r) + 1\,,
\end{align*}
where $Compl(r)$ is the complexity of the minimal derivation cost of representation $r$ according to the
LOT-grammar (see Table~\ref{tab:concepts}). Applied to our space of lexica, this construal assigns the highest probability 
to a lexicon of type $\Lall$, which only uses the simplest lexical representation ``all'' for
all messages. Lexica of type $\Llack$ are less likely, but more likely than $\Lbound$.

The menu of inductive biases argued to shape language acquisition is steadily being
refined. Apart from simplicity, prominent examples include mutual exclusivity
\citep{merriman+bowman:1989,clark:2009}, regularization \citep{hudson+etal:2005}, and
generalization \citep{smith:2011}. Even when these biases are considered in isolation, there
are many ways in which they can be translated into priors over types. The key assumption here,
common to all simplicity-biased priors, is that simple representational expressions should be
favored over more complex ones (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a,
  kirby+etal:2015}). In view of these numerous possibilities, we should stress that these
details -- from the generative grammar to its complexity measure -- are to be regarded as one
convenient operationalization of one general approach to explicating learning biases; this is
not a commitment that this general approach is necessarily superior or that, within it, this
particular instrumentalization is the single most plausible. We return to this issue in
Section~\ref{sec:discussion}.

\subsection{Simulation results}
\label{sec:simulation-results}
Recall that there are three parameters held constant across types: soft-max parameter $\lambda$
affects how strongly speakers favor messages that appear best from their subjective point of
view; the bottleneck size $k$ influences how faithfully learners can identify their teacher
type; $l$ defines the learners' disposition towards choosing the most likely teacher type from
the posterior distribution. Based on considerations and results from
Section~\ref{subsec:example}, we expect that competitor types (types with lexica of the kind
$\Lbound$) have a fitness advantage over target types (pragmatic types with lexica of the kind
$\Llack$), especially for very low levels of $\lambda$ (see Fig~\ref{fig:sketch}). Selection
based on fitness alone may therefore not lead to prevalence of target types in the
population. On the other hand, lexica of type $\Llack$ are simpler than those of type $\Lbound$
by the postulated measure from above. This may make them more likely to be adopted by learners,
especially when $l$ is high. Still, lexica of the kind $\Lall$ are in turn even more likely
\emph{a priori} than lexica of the kind $\Llack$. Simulation results will shed light on the
question whether target types can emerge, and for which parameter constellations.

To better understand the workings of a pressure towards communicative efficiency and a pressure
towards learnability, we look at the behavior of the replicator and mutator step first in
isolation, and then in combination. All simulation runs are initialized with an arbitrary
distribution over types, constituting a population's first generation. All reported results are
the outcome of 50 update steps. These outcomes correspond to developmental plateaus in which
change is, if not absent, then at least very slow. In other words, even if the resulting states
do not correspond to an eventual attracting state, they characterize population states in which the system remains for a long time. As specified in Section~\ref{sec:learnability}, the
mutation matrix $Q$ can be obtained by considering all possible state-message sequences of
length $k$. Given that this is intractable for large $k$, the sets of data which learners are
exposed to are approximated by sampling $250$ $k$-length sequences from each type's production
probabilities.

@@

Selection based on communicative success is sensitive to $\lambda$ since $\lambda$ influences
signaling behavior, which in turn determines communicative success. Fig. \ref{fig:only-R} shows
the proportion of target types, the highest competitor types and the highest type with $\Lall$
in a representative population after $50$ replicator steps. The plot also indicates the
proportion of the \emph{majority type}: the type with the highest proportion in the
population. With low $\lambda$ many types have very similar behavior, so that evolutionary
selection lacks grip and becomes very slow. The result is a very long transition with near
stagnancy in a rather homogeneous population with many types. Conversely, higher $\lambda$
promotes less-stochastic linguistic behavior, widening the gap in expressivity between some
types and promoting more homogeneous populations. As suggested by Fig. \ref{fig:only-R}, the
majority in most populations is not one of the six pragmatic $\Llack$-style types. That is, a
pressure only for communicative success does not lead to a prevalence of target types under any
$\lambda$-value. For instance, with $\lambda = 20$, $1000$ independent populations only had $11$
cases in which the target type was the majority type, corresponding to a mean proportion of
$.003$ across populations. By contrast, in $913$ cases the majority types had $\Lbound$ with
close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a
mean proportion of about $.48$ taken together. In sum, fitness-based selection of single types
requires sufficiently high $\lambda$ but does not often single out types with an underspecified
semantic representation. \mf{rewrite this paragraph or section}

@@

@@

@@

Pressure for communicative success and learnability are not sufficient on their own to have a
single target type dominate the population. When pressured for communicative success, the  communicative advantage of $\Lbound$ users leads to their prevalence. When pressured for
learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
complex alternatives such as $\Lbound$. However, learnability alone does not foment the
propagation of a single target type across the population.

@@

Combining pressures for communicative efficiency and learnability can lead to the selection of
a single target type (see Fig.~\ref{fig:rmd}). The proportion of a single majority target type
increases with $\lambda$ and $l$.

@@

Effects of manipulating the sequence length $k$ have not been addressed so far, but are rather
predictable: small values lead to more heterogeneous populations that reflect the learner's
prior more faithfully. This is due to the fact that the likelihood that a small sequence was
produced by any type is relatively uniform (modulo prior). By contrast, larger values
increasingly allow learners to differentiate types with different signaling behaviors.

@@

This case study should not be interpreted as a proposal for a definite explanation of how
scalar implicatures evolved. Other factors should be considered eventually even if they will
lead to much more complex modeling. One such factor is the observation that non-lexicalized
upper bounds allow a broader range of applicability. For example, when the speaker is not
certain as to whether \emph{all} is true. This may suggest an alternative and purely
functionalist argument for why upper-bounded meanings do not conventionalize: should contextual
cues provide enough information to the hearer to identify whether a bound is intended to be
conveyed pragmatically, then this is preferred over expressing it overtly through longer
expressions. For example, by saying {\em some but not all} explicitly. Importantly, although
morphosyntactic disambiguation may be dispreferred due to its relative length and complexity
\citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override
contextual cues that might otherwise mislead the hearer. In a nutshell, this explanation posits
that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to
communicate as economically as possible and pragmatic reasoning enables them to do so. What
this alternative argument does not explain is why functional pressure does not lead to the
emergence of different, equally costly lexical items to express different knowledge states of
the speaker (\citealt[252-267]{horn:1984},
\citealt{horn:1972,traugott:2004,vdAuwera:2010}). That is, this argument does not explain why
English and other languages do not have a monomorphemic dual for, e.g., {\em some} that
lexicalizes an upper-bound. If this hypothetical expression existed, it could be deployed to
signal that the speaker knows that {\em some but not all} holds, and unbounded {\em some} could
exclusively signal epistemic uncertainty. Looking at pressure from learnability might come in
again.

@@

\section{Conclusion}
The cultural evolution of meaning is influenced by intertwined pressures. We set out to
investigate this process by putting forward a model that combines pressure toward successful
information transfer with perturbations that may arise in the transmission of linguistic
knowledge in acquisition. Its objects of selection and replication are pairs of lexical
meanings and patterns of language use. This allows the model to trace the evolutionary interaction between
conventional meaning and pragmatic use. Additionally, it takes the challenge seriously of
neither semantics nor pragmatics being directly observable. Instead, learners need to infer
these unobservables from overt data that results from their combination.  These components and
their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in
weak scalar expression. This study showed that, when pressured for learnability and communicative
success, the former force can drive for simpler semantic representations inasmuch as pragmatics
can compensate for lack of expressivity in use. That is, the relative learning advantage of
simpler semantics in combination with functional pressure in use may offer an answer to why
natural languages fail to lexicalize systematic pragmatic inferences. And, more broadly, lead to a
division of labor between semantics and pragmatics.

>>>>

This section addresses the question under which conditions the division of labor, which was
described in Section~\ref{sec:introduction}, between underspecified semantics and pragmatic
enrichment could have evolved. We consider what are perhaps two of the simplest non-trivial
setups that speak to this matter and reflect on their limitations in
Section~\ref{sec:discussion}. Section \ref{subsec:example} first investigates a minimal example
intended to isolate effects of replication and mutation and to gain insights into the sense in
which semantic meaning and pragmatic use could be said to co-evolve. We then turn to a larger and
more realistic setup in
Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
for this case and their results.

@@

This section illustrates how replication and mutation act on the evolution of semantic meaning
and pragmatic use in the simplest non-trivial type space, which consists of only the four types
from Section~\ref{sec:languages+use}: a type has either lexicon $\Llack$ or $\Lbound$ and it is
either a literal (level-$0$) or a pragmatic (level-$1$) language user. We may think of such a
population as a point in a two-dimensional space. The grey dotted square in
Fig.~\ref{fig:sketch} outlines this space. The space's first dimension determines the
proportion of types that lexicalize ambiguous $\msome$ (the $y$-axis in
Fig.~\ref{fig:sketch}). The second dimension determines the proportion of pragmatic language
users (the $x$-axis in Fig.~\ref{fig:sketch}). Populations at the corners of this space are
monomorphic; all agents in them are of a single type. Points away from the corners represent
mixed populations of various types.\footnote{This two-dimensional representation underspecifies
  the exact proportions of types found in a mixed population. The space's center, for instance,
  could be composed in equal parts of pragmatic $\Llack$ and of literal $\Lbound$. But it could
  equally well be a population that has all four types in equal proportions. This level of
  granularity will nevertheless be sufficient for our purposes, as we are only interested in
  sketching rough trajectories after replicator or mutator steps.} We can predict general
trends in evolutionary trajectories in this space based on what we know from the literature and
the setup from Section~\ref{sec:fitn--learn}, even before any further simulation results.

@@

Despite its simplicity, the diagram in Figure~\ref{fig:sketch} makes clear the sense in which
semantics and pragmatics might co-evolve. Pragmatic language use will not evolve under a strong
and fully expressive semantics (bottom part of Figure~\ref{fig:sketch}). A weak semantic
convention will not evolve without disposition towards pragmatic enrichments (left part of
Figure~\ref{fig:sketch}). But together a weak semantics and a pragmatic disposition for
enrichment can co-evolve. These abstract theoretical considerations also suggest that we need
both fitness-based selection of communicative efficiency and pressure towards learnability to
see underspecified semantics and pragmatic language use emerge together.

@@

Fig.~\ref{fig:quiver} shows actual evolutionary trajectories in this type space. The replicator
steps shown in the top row of Fig.~\ref{fig:quiver} make the functional disadvantage of
$\Llack$ against $\Lbound$ apparent. With increasing $\lambda$, this difference pragmatic users
of $\Lbound$ are no longer selected against by fitness-based selection.

As for the iterated learning only, the diagrams in the bottom row of Fig.~\ref{fig:quiver} show
the case of a prior which favors types with underspecified semantics ($\Llack$) over those with
lexicalized upper bounds ($\Lbound$) by a factor of $1.05$. There are three main things to
note. First, as expected from the literature, a small bias for a type will lead to larger
proportions of types who adopt it under $l=1$ (see lower left space in Figure~\ref{fig:quiver},
where the system gravitates indeed to a population where the relative frequency of $\Llack$
over $\Lbound$ is exactly $1.05$). Second, if $l>0$, the dynamic is influenced heavily also by
asymmetries in production likelihoods. Third, these asymmetries in the production likelihoods
are in turn influenced by $\lambda$, so that $\lambda$ also has an effect on iterated Bayesian
learning (compare the middle and right-most diagram in the bottom row of
Fig.~\ref{fig:quiver}).

Fig.s~\ref{fig:sketch} and \ref{fig:quiver} investigate replicator and mutator steps in
isolation. It remains to investigate how exactly these forces shape a population over time when
they apply together. This will also hinge on which types are present because the space of all
types determines how much production likelihoods asymmetrically favor certain types over others
(when $l>1$; see Appendix~\ref{sec:example-iter-bayes}). For this reason, we turn to a somewhat
more realistic but still manageable setup next with a larger set of types.

@@

A lexicon $L$ is a mapping $\Messgs \rightarrow R$ from messages to representations. With three
messages there are $6^3 = 216$ possible lexica. Some assign the same representations to more
than one message and others lexicalize the same representations but associate them with
different messages. Out of these possible lexica, three kinds are of particular
relevance. First, lexica that assign the same lexical representations to more than one
message. Such lexica lack in expressivity but may be favored by particular learning biases
nonetheless (see below). Second, lexica that conventionalize upper-bounds to realize a
one-to-one mapping of messages to states. Finally, lexica that do not lexicalize upper bounds
but would allow for perfect communication under additional pragmatic strengthening. There are
six lexica of the second kind and six of the third.  The following three lexica exemplify each
kind:
\begin{align*}
  \begin{blockarray}{lccc}
    & & \underline{\Lall} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
     \mygray{\snone}  & 0 & 0 & 0\\
     \mygray{\ssome}  & 0 & 0 & 0\\
    \mygray{\sall}   & 1 & 1 & 1\\
    \end{block}
% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
  \end{blockarray} & &
 \begin{blockarray}{lccc}
    & & \underline{\Lbound} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 0 & 1\\
    \end{block}
%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
  \end{blockarray} & &
  \begin{blockarray}{lccc}
    & & \underline{\Llack} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 1 & 1\\
    \end{block}
%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
  \end{blockarray}
\end{align*}
% Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
% be equally likely; no information about the observed world state will be conveyed
% by its users. In contrast, users of $\Lbound$ can communicate world states perfectly, no matter
% whether they are literal or pragmatic users. Users of $\Llack$ can also communicate information
% about the actual world state but need pragmatic language use to approximate a one-to-one
% mapping between message use and states (see Section~\ref{sec:languages+use}).

@@

Note that while different types may lexicalize the same representations, they may nevertheless
map different states to different overt messages. Consequently, different types of the same
kind can fail to understand each other completely.

@@

The complexity measures for lexical representations from Table~\ref{tab:concepts} are used to
define a learning bias that favors simpler representations over more complex ones. The prior
probability of a type is just the prior probability of its lexicon. The prior of a lexicon is a
function of the complexity of the lexical representations in its image set. Lexica with simpler
representations accordingly have a higher prior. One simple way of defining such priors over
lexica (and thereby types) is:
\begin{align*}
  P(L)  & \propto \prod_{r \in Im(L)} P(r)   \ \text{, with} & 
  P(r) & \propto \max_{r'}Compl(r') - Compl(r) + 1\,,
\end{align*}
where $Compl(r)$ is the complexity of the minimal derivation cost of representation $r$ according to the
LOT-grammar (see Table~\ref{tab:concepts}). Applied to our space of lexica, this construal assigns the highest probability 
to a lexicon of type $\Lall$, which only uses the simplest lexical representation ``all'' for
all messages. Lexica of type $\Llack$ are less likely, but more likely than $\Lbound$.

The menu of inductive biases argued to shape language acquisition is steadily being
refined. Apart from simplicity, prominent examples include mutual exclusivity
\citep{merriman+bowman:1989,clark:2009}, regularization \citep{hudson+etal:2005}, and
generalization \citep{smith:2011}. Even when these biases are considered in isolation, there
are many ways in which they can be translated into priors over types. The key assumption here,
common to all simplicity-biased priors, is that simple representational expressions should be
favored over more complex ones (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a,
  kirby+etal:2015}). In view of these numerous possibilities, we should stress that these
details -- from the generative grammar to its complexity measure -- are to be regarded as one
convenient operationalization of one general approach to explicating learning biases; this is
not a implicit commitment to the claim that this particular instrumentalization is the single
most plausible. We merely want to be concrete to have a maximally concrete working example in
which the priors over types are systematically related to the complexity of lexical
representations. We return to this issue in Section~\ref{sec:discussion}.

\subsection{Simulation results}
\label{sec:simulation-results}
Recall that there are three parameters held constant across types: soft-max parameter $\lambda$
affects how strongly speakers favor messages that appear best from their subjective point of
view; the bottleneck size $k$ influences how faithfully learners can identify their teacher
type; $l$ defines the learners' disposition towards choosing the most likely teacher type from
the posterior distribution. Based on considerations and results from
Section~\ref{subsec:example}, we expect that competitor types (types with lexica of the kind
$\Lbound$) have a fitness advantage over target types (pragmatic types with lexica of the kind
$\Llack$), especially for very low levels of $\lambda$ (see Fig~\ref{fig:sketch}). Selection
based on fitness alone may therefore not lead to prevalence of target types in the
population. On the other hand, lexica of type $\Llack$ are simpler than those of type $\Lbound$
by the postulated measure from above. This may make them more likely to be adopted by learners,
especially when $l$ is high. Still, lexica of the kind $\Lall$ are in turn even more likely
\emph{a priori} than lexica of the kind $\Llack$. Simulation results will shed light on the
question whether target types can emerge, and for which parameter constellations this is
likely.

As before, we first look at the behavior of the replicator and mutator step in isolation, and
then in combination. All simulation runs are initialized with an arbitrary distribution over
types, constituting a population's first generation. All reported results are the outcome of 50
update steps. These outcomes correspond to developmental plateaus in which change is, if not
absent, then at least very slow. In other words, even if the resulting states do not correspond
to an eventual attracting state, they characterize population states in which the system
remains for a long time. As specified in Section~\ref{sec:learnability}, the mutation matrix
$Q$ can be obtained by considering all possible state-message sequences of length $k$. Given
that this is intractable for large $k$, the sets of data which learners are exposed to are
approximated by sampling $250$ $k$-length sequences from each type's production probabilities.

@@

Selection based on communicative success is sensitive to $\lambda$ since $\lambda$ influences
signaling behavior, which in turn determines communicative success. Fig. \ref{fig:only-R} shows
the proportion of target types, the highest competitor types and the highest type with $\Lall$
in a representative population after $50$ replicator steps. The plot also indicates the
proportion of the \emph{majority type}: the type with the highest proportion in the
population. With low $\lambda$ many types have very similar behavior, so that evolutionary
selection lacks grip and becomes very slow. The result is a very long transition with near
stagnancy in a rather homogeneous population with many types. Conversely, higher $\lambda$
promotes less-stochastic linguistic behavior, widening the gap in expressivity between some
types and promoting more homogeneous populations. As suggested by Fig. \ref{fig:only-R}, the
majority in most populations is not one of the six pragmatic $\Llack$-style types. That is, a
pressure only for communicative success does not lead to a prevalence of target types under any
$\lambda$-value. For instance, with $\lambda = 20$, $1000$ independent populations only had $11$
cases in which the target type was the majority type, corresponding to a mean proportion of
$.003$ across populations. By contrast, in $913$ cases the majority types had $\Lbound$ with
close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a
mean proportion of about $.48$ taken together. In sum, fitness-based selection of single types
requires sufficiently high $\lambda$ but does not often single out types with an underspecified
semantic representation.

@@

\mf{Possibly change plot on the left in Figure~\ref{fig:only-M} and adapt text
  accordingly. Write that we know that happens under $l = 5$ in the long run (even if we do not
  see it after 50 steps yet, if that is indeed so). If it is so, we could leave the plot and
  say: long term development is nice, but when fitness-based selection interferes at every step
  we do not care so much for infinite time, but rather for short time (actually one-step). If
  it is not so, insert a plot for $l>1$ that shows something interesting.}

@@

In sum, when pressured for learnability, pragmatic $\Llack$ is promoted over functionally
similar but semantically more complex alternatives such as $\Lbound$. However, learnability
alone does not foment the propagation of a single target type across the population, because it
does not differentiate between different ways of mapping the same semantic representations onto
different overt signals.

@@

@@

Pressure for communicative success and learnability are not sufficient on their own to have a
single target type dominate the population. But the combination of both pressures can lead to
the selection of a single target type (see Fig.~\ref{fig:rmd}). The proportion of a single
majority target type increases with $\lambda$ and $l$.

@@

Effects of manipulating the sequence length $k$ have not been addressed so far, but are rather
predictable: small values lead to more heterogeneous populations that reflect the learner's
prior more faithfully. This is due to the fact that the likelihood that a small sequence was
produced by any type is relatively uniform. By contrast, larger values increasingly allow
learners to differentiate types with different signaling behaviors.

@@

This case study is a first attempt at an explanation of how scalar implicatures evolved. But
other factors, which are presently not taken into account, should be considered eventually even
if they will lead to much more complex modeling. One such factor is the observation that
non-lexicalized upper bounds allow a broader range of applicability, e.g., when the speaker is
not certain as to whether \emph{all} is true. This may suggest an alternative and purely
functionalist argument for why upper-bounded meanings do not conventionalize: should contextual
cues provide enough information to the hearer to identify whether a bound is intended to be
conveyed pragmatically, then this is preferred over expressing it overtly through longer
expressions, e.g., by saying {\em some but not all} explicitly. Importantly, although
morphosyntactic disambiguation may be dispreferred due to its relative length and complexity
\citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override
contextual cues that might otherwise mislead the hearer. In a nutshell, this explanation posits
that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to
communicate as economically as possible and pragmatic reasoning enables them to do so. What
this alternative argument does not explain is why functional pressure does not lead to the
emergence of different, equally costly lexical items to express different knowledge states of
the speaker (\citealt[252-267]{horn:1984},
\citealt{horn:1972,traugott:2004,vdAuwera:2010}). That is, this argument does not explain why
English and other languages do not have a monomorphemic dual for, e.g., {\em some} that
lexicalizes an upper-bound. If this hypothetical expression existed, it could be deployed to
signal that the speaker knows that {\em some but not all} holds, and unbounded {\em some} could
exclusively signal epistemic uncertainty. Looking at pressure from learnability might come in
again.

@@

\section{Conclusion}
The cultural evolution of meaning is influenced by intertwined pressures. We set out to
investigate this process by putting forward a model that combines pressure toward successful
information transfer with perturbations that may arise in the transmission of linguistic
knowledge in acquisition. Its objects of selection and replication are pairs of lexical
meanings and patterns of language use. This allows the model to trace the evolutionary
interaction between conventional meaning and pragmatic use. Additionally, it takes the
challenge seriously of neither semantics nor pragmatics being directly observable. Instead,
learners need to infer these unobservables from overt data that results from their combination.
These components and their mutual influence were highlighted in a case study on the lack of
lexical upper-bounds in weak scalar expression. This study showed that, when pressured for
learnability and communicative success, the former force can drive for simpler semantic
representations inasmuch as pragmatics can compensate for lack of expressivity in use. That is,
the relative learning advantage of simpler semantics in combination with functional pressure in
use may offer an answer to why natural languages fail to lexicalize systematic pragmatic
inferences. Conversely, by appealing to economy of lexical representations, this model also
suggests a rationale for why a system of pragmatic enrichment could have evolved in the first
place. The resulting picture is one of co-evolution: the division of labor between semantics
and pragmatics could have evolved because underspecified lexical meanings are easier, yet can
only be maintained with a mechanism of pragmatic enrichment; pragmatic reasoning, however,
would not evolve in the absence of lexically ambiguous representations.
